\LARGE \bf Preparation of Papers for IEEE Sponsored Conferences \& Symposia*

### 2017-02-08 pwarel.tex
.

\newcommand{\pwa}{\rho}
\newcommand{\map}{f}
\newcommand{\amap}{f}
\newcommand{\gm}{\scr{T}}
\newcommand{\guard}{g}
\newcommand{\ds}{\scr{D}}

\chapter{Relational Modeling for Falsification}
\label{chap:rel}

In \chapref{abs}, we analyzed systems by restricting ourselves to
their black box semantics. This enabled us to reason about the
state-space reachability of the system without a direct analysis of
its structure. Using a coarse abstraction which we searched
on-the-fly, we could observe the local dynamics as required. This gave
us an efficient procedure to find abstract counter-examples.  However,
to concretize the counter-examples, a `closer look' or refinement of
the abstraction is required. The grid based state-space abstraction
was refined by splitting the relevant cells (abstract states) into
smaller ones. As noted previously, due to the curse of dimensionality,
such a uniform splitting is an expensive operation, and can not scale to
higher dimensions.

In this chapter, we further our exploration of trajectory segment
based methods; and explore an alternative approach to overcome the
explosion in abstract states. Instead of selectively refining the
abstraction, we compute a model of the black box system and use
bounded model checking to find a concrete counter-example in the
model~\footnote{Due to modeling errors, this might not be reproducible in
the original black-box system.}. We then use this counter-example to
guide the search towards a counter-example in the original system.

More specifically, we use regression to quantitatively estimate the
discovered relations, which were witnessed by trajectory segments, by
affine maps. These maps approximate locally observed behaviors, which
are incorporated into the sampled reachability graph as edge
annotations.  The resulting graph or the Piece-Wise Affine (PWA)
relational model can be interpreted as an infinite state discrete
transition system and model checked for time bounded safety
properties. A counter-example if found, can indicate the presence of a
violation in the system.
%Furthermore, using linear programming, we can over-approximate the
%neighbhourhood of the counter-example in the model.

Towards the end of the chapter, we discuss extensions to data driven
approaches (instead of simulator driven), where instead of a
simulator, a fixed set of data is provided as the behavioral
description of the system. Using a combination of relational modeling,
and program analysis we outline the future work for falsifying
properties of SDCS.

\section{Overview}
\label{sec:ovw}

@@

\input{figs/relational/segtraj_rel_enriched.tex}

@@

\section{Background}

In this section, we present the basics of relational abstractions and
fundamentals of regression analysis.

\subsection{Relational Abstraction}

For general hybrid dynamical systems, algorithmic ways for computing
analytical solutions do not exist. Discretization (along with suitable
abstractions) is often employed to transform the systems into a
discrete transition system. Relational abstractions is one such idea,
which abstracts continuous dynamics by discrete relations using
appropriate \textit{reachability invariants}. Both time independent
and time dependent relations have been proposed; the former captures
all reachable states over all time, whereas, the latter explicitly
includes time by relating reachable states to time. Thus it
can prove timing properties whereas, time independent relational
abstractions can not. In both cases, the resulting abstractions can be
interpreted as discrete transition systems and can be analyzed using
model checkers. We briefly discuss these two types of abstractions.

\mypara{Timeless Relational Abstraction}

Relational abstractions for hybrid systems were proposed in
\cite{Sankaranarayanan+Tiwari/2011/Relational}. For a hybrid automaton
model, they can summarize the continuous dynamics of each mode using a
binary relation over continuous states. The resulting relations are
timeless (independent) and hence valid for all time as long as
the mode invariant is satisfied. The relations take the general form
of $R(\x,\x') \bowtie 0$, where $\bowtie$ represents one of the
relational operators $=, \ge, \le, <, >$. For example, an abstraction
that captures the monotonicity with respect to time for the
differential equation $\dot{x} = 2$ is $x' > x$. The abstraction
capturing the relation between the set of ODEs: $\dot{x} = 2$,
$\dot{y} = 5$, is $5(x' - x) = 2(y' - y)$.

Such relation summaries, discretize the continuous evolution of a
given system into an infinite state transition system, in which safety
properties can be verified using $k$-induction, or falsified using
bounded model checkers. The relationalization procedure involves
finding suitable invariants in a chosen abstract domain, such as
affine abstractions, eigen abstractions or box abstractions
\cite{Sankaranarayanan+Tiwari/2011/Relational}.  For this, different
techniques like template based invariant generation
\cite{Gulwani+Tiwari/2008/Constraint,
Colon+Sankaranarayanan+Sipma/03/Linear} can be used.

\mypara{Timed and Time-Aware Relational Abstractions}

As the above discussed relations are timeless, we cannot reason over
the timing properties of the original system. Moreover, they are not
suitable for time-triggered systems. Timed relational
abstractions~\cite{zutshi2012timed} and time-aware relational
abstractions~\cite{mover2013time} were proposed to overcome these
shortcomings by explicitly including time in the relations. In
addition, both can be more precise than their timeless counterpart,
but unlike it, assume continuous affine dynamics.

To analyze time-triggered systems like an SDCS modeled by an affine
hybrid automata~\footnote{An affine hybrid automata is restricted to
the ODEs, resets and guards being affine.}, timed relational
abstractions can be computed directly from the solutions of the
underlying affine ODEs.  Recall that the solution of an affine ODE
$\dot{\x} = A\x$ can be represented using the matrix exponential as
$\x(t) = e^{tA}\x(0)$. This gives us the timed relation $\x' =
e^{tA}\x$.  Clearly, the relation is non-linear with respect to time.
However, for the case of SDCS with a fixed sampling time period, we
obtain linear relations in $\x$.  We demonstrated the usefulness of
timed relational abstractions for the case of linear systems
in \cite{zutshi2012timed}, and now incorporate the idea to discretize
black box dynamical systems.

Similar to timeless relational abstractions, time-aware relational
abstractions \cite{mover2013time} construct binary relations between
the current state $\x$ of the system and any future reachable state
$\x'$. The difference lies with the latter also constructing relations
between the current time $t$ and any future time $t'$. This is
achieved by a case by case analysis of the eigen structure of the
matrix $A$ (of an affine ODE). Separate abstractions are used for the
case of linear systems with constant rate, real eigen values and complex eigenvalues.

\subsection{Learning Dynamics using Simple Linear Regression}

\mypara{Regression Analysis}

In statistics, regression is the problem of finding a
\textit{predictor}, which can suitably predict the relationship
between the given set of observed input $\x$ and output $\y$ vectors.
In other words, assuming that $\y$ depends on $\x$, regression
strategies find either a parameterized or a non-parameterized
prediction function to explain the dependence. We now discuss simple
linear regression, which is parametric in nature and searches for an
affine predictor. It is also called \textit{ordinary least squares}.

\mypara{Ordinary Least Squares (OLS)}

Let the data set be comprised of $N$ input and output pairs $(\x,
y)$, where $\x\in\reals^n$ and $y\in\reals$. If $N>n$, which is
the case in the current context of \textit{finding the best fit}, the
problem is over-determined; there are more equations than
unknowns. Hence, a single affine function cannot be found which
satisfies the equation $\forall i\in\{1\ldots N\}. y_i = \vec{a}^T\x_i + b$. Instead, we
need to find the `best' choice for $\vec{a}$ and $b$. This is formally
defined using a loss function. For the case of simple linear
regression or OLS, the loss function is the sum of squares of the
errors in prediction.  The task is then to determine the vector of
coefficients $\vec{a}$ and an offset constant $b$, such that the least
square error of the affine predictor is minimized for the given data
set.
\begin{equation}
    \operatornamewithlimits{argmin}_{\vec{a}, b}\displaystyle\sum_{i=1}^{N}{\left(\y_i - (\vec{a}^T\x_i + b)\right)^2}
\label{reg}
\end{equation}
To ease the presentation, we can rewrite the above as a homogeneous
expression by augmenting $\x$ by $\hat{\x} = \begin{bmatrix}\x \\ 1\end{bmatrix}$ and replacing $\vec{a}$ and $b$ by the vector
$\vec{ab} = \begin{bmatrix}\vec{a} \\ b \end{bmatrix}$. The equation~\ref{reg} now becomes
\[
    \operatornamewithlimits{argmin}_{\vec{ab}}\displaystyle\sum_{i=1}^{N}{(\y_i - \vec{ab}^T \hat{\x}_i)^2}
\]
The solution of OLS can be analytically computed as
\[ Ab = (X^TX)^{-1}X^T\y\]
where $X$ is the matrix representing the horizontal stacking of all
$\hat{\x}$.
The details can be found in several texts on learning and statistics
\cite{friedman2001elements}.

Given a time invariant dynamical system $\x' = \simulate(\x, \Delta)$,
where $\x\in\reals^n$ we can use OLS to approximate its trajectories
at fixed time step $\Delta$ by a discrete map $\x' = A\x + \vb$, where
$A = \begin{bmatrix}\vec{a}_1^T \\ \vdots \\ \vec{a}_n^T \end{bmatrix}$ and $\vb =
\begin{bmatrix}b_1 \\ \vdots \\ b_n \end{bmatrix}$. This map is a
    global relational model for the system. The data set for the
    regression is a set of pairs $\setof{(\x,\x') | \x' =
    \simulate(\x, \Delta)}$ and can be generated on demand. Another
    set can be generated to compute the error $\delta$ as an interval
    of vectors, where each element is an interval $\delta_i \in
    [\delta_{min}, \delta_{max}]$.

    Affine maps are poor
    approximations for arbitrary non-linear functions. Hence, we use a
    collection of affine maps to approximate the local behaviors (in
    state-space) of the system $\simulate$. This results in a piece
    wise approximation, as described in the next section.

\subsection{Piecewise Affine (PWA) Transition System}

We define the PWA transition system as a transition system
(ref.~\secref{DiscSys}) $\rho:\tupleof{\pLocs,\pVars,\scrT, \pLocI,\Theta}$
where each transition $\tau \in \scrT$, is associated with a
transition relation of the form
\[
    \rho_{\tau}(\pVar,\pVar') \subseteq \setof{(\pVar,f_{\tau}(\pVar)) \;|\; g_\tau(\pVar) \land g'_\tau(f_{\tau}(\pVar))}
\]
where $f_{\tau}$ is an affine map relating the states and $g_{\tau}$,
$g'_{\tau}$ are affine guards (pre and post conditions) on the states.
The assertion on the initial states $\Theta$ is also affine.

We can now use a PWA transition system $\rho$ to approximate the
behavior of a hybrid dynamical system $\System$ defined by a simulator
$\simulate_{\System}$~\footnote{Due to $f_\tau$ being restricted to an affine
form, we can only approximate general hybrid systems.}. Each
transition of $\rho$ represents a discrete time step $\Delta$. Let the
state-space of $\System$ be given by $\x\in\HybridStates^n$, then the
states of $\rho$ are also given by $\x\in\HybridStates^n$. The affine guard
predicates $g_\tau$ are given by a conjunction of hyper-planes and
hence can be represented in matrix form as a polyhedron (defined by
$m$ constraints) in the state-space
\[
g:C\x - \vd \le 0
\]
where $C$ is an $m \times n$ matrix and $\vd$ is a vector of length
$m$. The affine maps $f_\tau$ can be represented in matrix form as
$f:\x \mapsto A\x+\vb$. To make $f_\tau$ `richer', we incorporate
an error term $\delta$ which is defined by a vector of intervals
$[\delta_{\min},\delta_{\max}]$. Hence,
\[
f:\x \mapsto A\x + \vb + \delta
\]
defines a set valued relation of the form $R:\setof{(\x,\x') \;|\; \x'
\in f_\tau(\x)}$.

% Given a state vector $\x\in\reals^n$, a guarded affine map
% $\gm:(\guard, \amap)$ defines the discretized consecution rule as a
% pair of an affine guard predicate $\guard:C\x - \vd \le 0$ and an affine map
% $\amap:A\x+\vb$, where $A \in \reals^{n \times n}$, $C \in
% \reals^{m \times n}$ are matrices and $\vb \in \reals^n$, $\vd
% \in \reals^m$ are vectors. A guarded affine map is satisfied if its
% guard is satisfied.

We now formalize the PWA affine transition system for a dynamical system.
\begin{definition}[PWA Transition System]

Given a hybrid dynamical system over a state-space
$\HybridStates^n$, a PWA transition system $\rho$ is given by the
tuple $\tupleof{\pLocs,\x,\scrT,\pLocI,\Theta}$, where $\tau \in
\scrT$ are affine transition relations and $\Theta$ is an affine
predicate over $\x$ and the states $\x\in\HybridStates^n$. The
transition relation is then defined by $\scrT$ with $n$ transition
relations as follows

\begin{equation}
    \scrT = \left\{
        \begin{array}{ll}
            g_1(\x) \land g'_1(\x') \implies \x' \in f_1(\x) \\
            \ldots\\
            g_n(\x) \land g'_n(\x') \implies \x' \in f_n(\x) \\
        \end{array}
    \right.
\end{equation}

\end{definition}

where $f_i:\x \mapsto A_i\x + \vb_i + \delta_i$,
$g_i(\x):C_i\x-\vd_i\le0$, $g'_i(\x'):C'_i\x'-\vd'_i\le0$ and $\x'$
denotes the next state of the system.

A PWA model is \textit{deterministic}, iff for every state
$\x\in\HybridStates^n$, a unique guarded affine map is satisfied and
all errors $\delta$ are singletons. However, in practice such a case
rarely exists. A PWA model will be usually non-deterministic, both due
to multiple choice of transition relations and the reachable states at
the $k^{th}$ time step (or after $k\Delta$ units of time) being a set.
Abusing the notation, we denote the set of states reachable in a
single step in $\rho$ by $\x' = \rho(\x)$.

A PWA transition system is \textit{complete}, iff for every state
$\x\in\HybridStates^n$, there exists at least one satisfied
transition relation $\scrT$. If this is not the case, the system can
deadlock, with no further executions. This usually results from
incorrect modeling of SDCS, and from here on, we do not consider such
cases.

% Given a dynamical system over a continuous state-space
% $\ContStates\in\reals^n$, a PWA model is a map $\ContStates \mapsto
% \setof{\gm_1, \gm_2, \ldots, \gm_n}$ from the state-space of the
% dynamical system to a finite set of guarded affine maps
% $\setof{\gm_1, \gm_2, \ldots, \gm_n}$. It defines the consecution
% rule by the affine map of the satisfied $\gm$.

% \begin{equation}
%     \pwa = \left\{
%         \begin{array}{ll}
%             \gm_1: \amap_1(\x) &\text{if}\; \guard_1(\x) \\
%             \gm_2: \amap_2(\x) &\text{if}\; \guard_2(\x) \\
%             \ldots & \ldots\\
%             \gm_n: \amap_n(\x) &\text{if}\; \guard_n(\x) \\
%         \end{array}
%     \right.
% \end{equation}

% \end{definition}

% where $\amap_i = A_i\x + \vb_i$, and $\guard_i(\x) \equiv C_i\x-\vd_i\le0$.
% Abusing the notation, we denote the next state computed by the PWA
% model as $\x' = \pwa(\x)$. A PWA model is \textit{deterministic}, iff
% for every state $\x\in\ContStates$, a unique guarded affine map is
% satisfied. A PWA model is \textit{complete}, iff for every state
% $\x\in\ContStates$, there exists at least one satisfied guarded affine
% map $\gm$.

\section{Relational Modeling}

We now describe how relational models can be computed for a given
black box system by enriching the abstract reachability graph obtained
by scatter-and-simulate. We first formalize the notion of an enriched
graph, and then show how they can be interpreted as a PWA transition
system using $k$-relational modeling.

\subsection{Abstract Enriched Graph}

The existential abstraction relation in \chapref{abs} was defined as
follows:
\[
%\exists \x\in\C. \exists \x'\in\C'. \x'=\simulate(\x, \tau) \iff \C\rel{\tau}\C'
    C \areach{t} C' \iff \exists \x \in C.\ \exists \x' \in C'.\ \x \areach{t} \x'
\]
The abstract relation $\areach{t}$ can be enriched by incorporating
the affine relation between $\x$ and $\x'$. For an arbitrary dynamical
system, such a relation can be rarely represented using an exact affine map.
This is due to the presence of non-linear and hybrid behaviors. But,
an affine map can always be estimated with an error.

If the system dynamics are completely specified in the form of a white
box model, we can use first order approximations to find the affine
expressions for the relations. Using a tool like \flowstar, we can
obtain sound over-approximate affine maps of the form $A\x + [\vb^l,
\vb^h]$, where $[\vb^l,\vb^h]$ denotes the interval of vectors, such
that, every element $b_i$ of the vector $\vb$ is contained in the
respective scalar interval $b_i\in[b^l_i,b^h_i]$.

For the case of black box systems, such a sound approximation is not
possible. Instead, we rely on a statistical method like simple linear
regression to estimate the affine map $\amap: A\x + \vb$. We then
estimate the error $\delta$ and generalize $\amap$ to an interval
affine map as before $\amap: A\x + \vb + \delta$.  Finally, we get an
abstraction with the below relation
\[
    C \areach{f} C' \iff \exists \x \in C.\ \exists \x' \in C'.\ \x' \in A\x + \vb \pm \delta.
    %C \areach{A\x+\vb+\delta} C' \iff \exists \x \in C.\ \exists \x' \in C'.\
\]

However, in the presence of complex non-linear behavior, using a
single affine map can lead to a poor approximation. To increase the
precision, one can use more than one affine map. Let us denote this set as
$R_{(C,C')}$ (ref.~\secref{ovw}).
\[
    R_{(C,C')}(\x,\x'): \setof{f_i \;|\; \exists \x \in C.\ \exists \x' \in C'.\ \x' \in f_i(\x)}
\]

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}

    \draw [line] (-1.2,-1.2) rectangle (1.2,1.2);
    \draw[arw] (-3,0) -- (-1.2,0);
    \draw[arw] (1.2,0) -- (3,0);

\node at (-3.4,0) {$G$};
\node at (3.4,0) {$G^R$};
\node[align=center] at (0, 0) {\scriptsize Enrich using \\ \scriptsize Regression \\ \scriptsize(OLS)};

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{Using OLS, $G^R$ is computed by determining the appropriate
$R_{(C,C')}$ for each edge of $G$.}
\label{fig:enrichment}
\vspace*{-.3cm}
\end{figure}

As shown in \figref{enrichment}, we use OLS to compute an enrichment
of the abstraction graph $G$. For every edge $(C,C') \in edges(G)$,
the set of relations $R_{(C,C')}(\x,\x')$ is computed. The semantics
of the enriched edge are clearly non-deterministic. Form a state $\x
\in C$, any $f_i\in R_{(C,C')}$ can be taken as long as $\x' \in C$'
is satisfied. This interpretation results in an infinite state
transition system. We now detail the construction of the set
$R_{(C,C')}$ using $k$-relational modeling.

% We now compute a PWA model for a given black box system by estimating
% the dynamics using OLS. We can estimate the affine relations between
% every cell, but as discussed in ~\chapref{abs}, it is a futile
% approach.  Instead, we use the same heuristic as before:
% scatter-and-simulate, to select the cell relations for estimation. We
% build the reachability graph as before, but in addition, annotate each
% edge with the values of estimated $A$, $\vb$, and $\delta$ for the
% respective relation. The reachability graph thus computed, is a
% transition system. Using off-the-shelf bounded model checkers, we
% reason about the system's safety properties.

\subsection{$k$ - Relational Modeling}

\begin{figure}[!htbp]
\vspace*{-.2cm}
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}

\begin{scope}[scale=1]
\draw [line] (-1.0,0) circle (0.5);
\draw [line] (-4.0,0) circle (0.5);
\draw [line] (2.0,1) circle (0.5);
\draw [line] (2.0,-1) circle (0.5);
\draw[arw] (-0.5,0) -- (1.5,1.0);
\draw[arw] (-0.5,0) -- (1.5,-1.0);
\draw[arw] (-3.5,0) -- (-1.5,0.0);
\node at (-4.0,0) {$C$};
\node at (-1.0,0) {$C'$};
\node at (2.0,1.0) {$C''_1$};
\node at (2.0,-1.0) {$C''_2$};
\draw [line] (2.0,-1) circle (0.5);
\draw[arw] (-5.0,-2.0) -- (3.0,-2.0);
\node at (-4.0,-2.5) {$k=0$};
\node at (-1.0,-2.5) {$k=1$};
\node at (+2.0,-2.5) {$k=2$};
\draw [fill=black] (-4.0,-2.0) circle (0.05);
\draw [fill=black] (-1.0,-2.0) circle (0.05);
\draw [fill=black] (2.0,-2.0) circle (0.05);

\end{scope}
\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{Nodes/Cells of $G$ shown along with increasing values of $k$.}
\label{fig:k-rel}
\vspace*{-.3cm}
\end{figure}

Given a $G$, when computing the set of relations for an edge $R$, we
consider only the set of abstract states reachable at a specific time
step.  Intuitively, one might consider the states reachable in one
time step $\Delta$.  However, such a process can be generalized by
looking at the states reachable at ${0\Delta, 1\Delta, \ldots,
k\Delta}$ time steps.

Using \figref{k-rel}, we illustrate this notion. To compute
$R_{(C,C')}$, we
split the data set $D(C,C')$ consisting of trajectory segments (of time lengths
$k\Delta$) as follows.  We observe the local behavior of the system by
noting the evolution of the system $\System$ from a state $\x(t)
\in C$ for a time length dependant on $k$. For
\begin{itemize}
    \item{$k=0$}: we observe all trajectory segments of length $\Delta$.
    \item{$k=1$}: we observe trajectory segments of length $\Delta$, which satisfy $\x(t+\Delta) \in C'$.
    \item{$k=2$}: we observe trajectory segments of length $2\Delta$,
        and split them in two sets (a) and (b) on the basis of the
        cells reached at time $(t+\Delta)$ and $(t+2\Delta)$.
        \begin{enumerate}[(a)]
            \item $\x(t+\Delta) \in C' \land \x(t+2\Delta) \in C''_1$
            \item $\x(t+\Delta) \in C' \land \x(t+2\Delta) \in C''_2$
        \end{enumerate}
    \item{$k=n$}: we observe trajectory segments of length $n\Delta$,
        and split them in to multiple sets on the basis of the cells
        reached at time $(t+\Delta),\; (t+2\Delta),\;
        \ldots,\;(t+n\Delta)$.
\end{itemize}

$k$-relational modeling can be understood as a heuristic, which uses
the underlying abstraction to differentiate behaviors of the system
which `diverge' (or are revealed to be `distinct' at a future time).
Such a heuristic can be useful in increasing the precision of the
learnt affine maps.  We now formalize this notion for $k=0$, $1$, and
$n$ and illustrate using examples.

\mypara{$0$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}[scale=1]

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    %\draw [line] (3.0,1.0) rectangle (5.0,3.0);
    %\draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);
    \draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (3.7,2.2) {\scriptsize{$\x_0'$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
\node at (3.7,1.7) {\scriptsize{$\x_1'$}};
\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.7,-1.2) {\scriptsize{$\x_2'$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\node at (1.7,-1) {$\pi_3$};
\draw[arw] (-1.5,-2.0) -- (4.5,-2.0);
\node at (0,-2.5) {$t=0$};
\node at (4,-2.5) {$t=\Delta$};
\draw [fill=black] (0.0,-2.0) circle (0.05);
\draw [fill=black] (4.0,-2.0) circle (0.05);

\end{scope}
\begin{scope}[xshift=0,yshift=3.0cm,scale=1]
    \node at (2,0) {\footnotesize $\scr{T} \equiv g_1(\x):\vx \in C
    \implies A_1\vx + \vb_1 + \delta_1$};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{All the trajectory segments will be used to construct the
model; $\ds = \setof{\pi_1, \pi_2, \pi_3}$.}
\label{fig:k0}
\vspace*{-.3cm}
\end{figure}

When $k=0$, the $0$-relational model is a PWA transition system which only
defines the evolution of states $\x$, and does not specify the
reachable cell. The guard predicates of the relations are defined over
cells: $g_i(\x): \x \in C_i$ while $g'_i(\x'): True$.

We use regression to estimate the dynamics for the outgoing trajectory
segments form a cell $C$. Hence, the data set $\ds$ for the regression
includes all trajectory segments beginning from the same cell
\[
    \ds = \setof{\pi_t | start(\pi_t) \in C}.
\]
This includes trajectory segments ending in different cells, as shown
in ~\figref{k0}. For $N$ cells, this results in a $\rho$ with $N$ transition relations.
\begin{equation*}
    \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C_1 \implies A_1\vx + \vb_1 + \delta_1\\
            \ldots \\
            g_n(\x):\vx \in C_n \implies A_n\vx + \vb_n + \delta_n\\
        \end{array}
    \right.
\end{equation*}

Note that this can be quite imprecise when the cells are big,
containing regions of state-space with complex dynamics. This is true
for both non-linear systems and hybrid dynamical system, where a cell
can contain two or more modes with differing continuous dynamics.

\mypara{$1$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}[scale=1]

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    \draw [line] (3.0,1.0) rectangle (5.0,3.0);
    \draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);
    \draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (4.0, 2.0) {$C'_1$};
\node at (4.0, -2.0) {$C'_2$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.3,2.5) {\scriptsize{$\x_0'$}};
\node at (3.7,1.5) {\scriptsize{$\x_1'$}};
\node at (3.7,-1.5) {\scriptsize{$\x_2'$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\node at (1.7,-1) {$\pi_3$};
\draw[arw] (-1.5,-3.5) -- (5.5,-3.5);
\node at (0,-4.0) {$t=0$};
\node at (4,-4.0) {$t=\Delta$};
\draw [fill=black] (0.0,-3.5) circle (0.05);
\draw [fill=black] (4.0,-3.5) circle (0.05);

\end{scope}
\begin{scope}[xshift=0,yshift=4.0cm,scale=1]
    \node at (2,0) {\begin{minipage}{\textwidth}
    \footnotesize \begin{equation*}
        \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
            g_2(\x):\vx \in C \land g'_2(\x):\vx \in C'_2 \implies A_2\vx + \vb_2 + \delta_2
        \end{array}
    \right.
\end{equation*}
\end{minipage}};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{The data gets split into two sets $\ds_1 = \setof{\pi_1,
    \pi_2}$ and $\ds_2 = \setof{\pi_3}$ and two relations:
    $R_{(C,C'_1)}$ and $R_{(C,C'_2)}$, each with one affine map, are constructed.}
%Cardinality($R_{(C,C')}$) = $2$.}
\label{fig:k1}
\vspace*{-.3cm}
\end{figure}

% \begin{scope}[xshift=0,yshift=4.0cm,scale=1]
%     \node at (2,0) {\footnotesize $\begin{equation*}
%     \scr{T} = \left\{
%         \begin{array}{ll}
%             g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
%             g_2(\x):\vx \in C \land g'_2(\x):\vx \in C'_2 \implies A_2\vx + \vb_2 + \delta_2
%         \end{array}
%     \right.
% \end{equation*}$};
% \end{scope}

To improve the preciseness of learnt dynamics, we include the
reachability relation in the
regression. For every relation $C\areach{t}C'$, the data set $\ds$ is
comprised only of trajectory segments $\pi_t$ which start and end in
the same set of cells.
\[
    \ds = \setof{\pi_t | start(\pi_t) \in C \land end(\pi_t) \in C'}.
\]
For $N$ edges in $G$, $1$-relationalization results in a $\rho$ with
$N$ transition relations.

\figref{k1} illustrates the $k=1$ refinement of the case shown in
\figref{k0}. The data set $\ds$ is split into two data sets $\ds_1
= \setof{\pi_1, \pi_2}$ and $\ds_2 = \setof{\pi_3}$, using which, two
relations are constructed. The resulting $1$-relational model
is at least as precise as the corresponding $0$-relational model.

\mypara{$k$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    \draw [line] (3.0,1.0) rectangle (5.0,3.0);
    %\draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (7.0,1.0) rectangle (9.0,3.0);
    \draw [line] (7.0,-3.0) rectangle (9.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (3.5,2.2) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (8.5,1.8) ;
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (3.5,1.7) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (8.5,-2.2) ;

%     \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
%     \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);

    %\draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (4.0, 2.0) {$C'_1$};
\node at (8.0, 2.0) {$C''_1$};
\node at (8.0, -2.0) {$C''_2$};
%\node at (4.0, -2.0) {$C'_2$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
%\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.3,2.5) {\scriptsize{$\x_0'$}};
\node at (3.7,1.5) {\scriptsize{$\x_1'$}};
%\node at (3.7,-1.5) {\scriptsize{$\x_2'$}};
\node at (8.7,1.6) {\scriptsize{$\x_0''$}};
\node at (8.5,-2.7) {\scriptsize{$\x_1''$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\draw[arw] (-1.5,-3.5) -- (9.5,-3.5);
\node at (0,-4.0) {$t=0$};
\node at (4,-4.0) {$t=\Delta$};
\node at (8,-4.0) {$t=2\Delta$};
\draw [fill=black] (0.0,-3.5) circle (0.05);
\draw [fill=black] (4.0,-3.5) circle (0.05);
\draw [fill=black] (8.0,-3.5) circle (0.05);
\end{scope}

\begin{scope}[xshift=0,yshift=4.0cm,scale=1]
    \node at (4,0) {\begin{minipage}{\textwidth}
        \footnotesize \begin{equation*}
    \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_2\vx + \vb_2 + \delta_2
        \end{array}
    \right.
\end{equation*}
\end{minipage}};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{The $k=2$ refinement further splits the data set $\ds_1$
    into two sets $\ds_{11} = \setof{\pi_1}$ and $\ds_{12} =
    \setof{\pi_2}$ and $R_{(C,C'_1)}$ now has two affine maps,
    non-deterministically defining the system behavior.}
\label{fig:k2}
\vspace*{-.3cm}
\end{figure}

Finally, we generalize the relational models to $k$-relational PWA
models. A $k$-relational model is constructed by using $k$ length
\textit{connected} segmented trajectories. A segmented trajectory $\SegTraj$ is
\textit{connected} iff its cost $\cost(\SegTraj) = 0$.

Two $\SegTraj:\tupleof{\pi_{t_1}, \ldots, \pi_{t_k}}$ and
$\SegTraj':\tupleof{\pi'_{t_1}, \ldots, \pi'_{t_k}}$ are
\textit{similar} if their trajectory segments have the same sequence
of cell traversals $\tupleof{C_1, C_2, \ldots, C_k}$, \ie
\[
    \forall i\in\setof{1 \ldots k}.
    start(\pi_{t_i}) \in C_i
    \land start(\pi'_{t_i}) \in C_i
    \land end(\pi_{t_i}) \in C_{i+1}
    \land end(\pi'_{t_i}) \in C_{i+1}
\]

>>>>

\subsection{Overview}
\label{sec:ovw}

@@

\input{figs/segtraj_rel_enriched.tex}

@@
